Type-Safe Modular Parsing

### 2016-10-15 paper/sections/Library/ObjectAlgebras.tex
More text.

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The recursive types are represented using generics. Here we show the example of a simple
expression language:
\begin{lstlisting}
trait ExpAlg[E] {
    def lit(i : Int) : E
    def add(e1 : E, e2 : E) : E
}
\end{lstlisting}
The language has two constructors: literals and additions. The trait \lstinline{ExpAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of evaluation:
\begin{lstlisting}
trait Eval extends ExpAlg[Int] {
    def lit(i : Int) = i
    def add(e1 : Int, e2 : Int) = e1 + e2
}
\end{lstlisting}
Here \lstinline{Eval} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns an integer as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:
\begin{lstlisting}
trait Print extends ExpAlg[String] {
    def lit(i : Int) = "" + i
    def add(e1 : String, e2 : String) = e1 + " + " + e2
}
\end{lstlisting}
On the other hand, the data variants can be extended by inheriting \lstinline{ExpAlg} and adding new cases only. Suppose we want to
have subtraction in expressions, a new Object Algebra interface \lstinline{SubAlg} is defined as follows:
\begin{lstlisting}
trait SubAlg[E] extends ExpAlg[E] {
    def sub(e1 : E, e2 : E) : E
}
\end{lstlisting}
Now evaluation on the new language can be realized by code reuse, and without modifying existing code:
\begin{lstlisting}
trait SubEval extends Eval {
    def sub(e1 : Int, e2 : Int) = e1 - e2
}
\end{lstlisting}
On this, to create an expression of \lstinline{SubAlg}, a generic method is defined as follows:
\begin{lstlisting}
def build[E](alg : SubAlg[E]) : E =
    alg.sub(alg.add(alg.lit(2), alg.lit(3)), alg.lit(4))
\end{lstlisting}
Such a method implicitly represents the expression \lstinline{"2 + 3 - 4"}. The code
\begin{lstlisting}
build(new Eval(){})
\end{lstlisting}
results in \lstinline{1}, as the result of evaluation. Now we observe that an ``object'' of expression actually
has the function type \lstinline{SubAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not a case, because functions can be used
in a very clever way. For example, if one wants to refactor an \lstinline{ExpAlg} expression to a \lstinline{SubAlg} expression by
replacing all negative numbers with subtraction, a transformation algebra can be implemented as follows:
\begin{lstlisting}
trait Refactor[E] extends ExpAlg[E] {
    def alg : SubAlg[E]
    def lit(i : Int) = if (i > 0) alg.lit(i) else alg.sub(alg.lit(0), alg.lit(-i))
    def add(e1 : E, e2 : E) = alg.add(e1, e2)
}
\end{lstlisting}
With \lstinline{Refactor}, the expression \lstinline{"1 + (-2)"} is refactored to \lstinline{"1 + (0 - 2)"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply refactoring to an \lstinline{ExpAlg} expression.

>>>>

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The recursive types are represented using generics. Here we show the example of a simple
expression language:
\begin{lstlisting}
trait ExpAlg[E] {
    def lit(i : Int) : E
    def add(e1 : E, e2 : E) : E
}
\end{lstlisting}
The language has two constructors: literals and additions. The trait \lstinline{ExpAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of evaluation:
\begin{lstlisting}
trait Eval extends ExpAlg[Int] {
    def lit(i : Int) = i
    def add(e1 : Int, e2 : Int) = e1 + e2
}
\end{lstlisting}
Here \lstinline{Eval} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns an integer as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:
\begin{lstlisting}
trait Print extends ExpAlg[String] {
    def lit(i : Int) = "" + i
    def add(e1 : String, e2 : String) = e1 + " + " + e2
}
\end{lstlisting}
On the other hand, the data variants can be extended by inheriting \lstinline{ExpAlg} and adding new cases only. Suppose we want to
have subtraction in expressions, a new Object Algebra interface \lstinline{SubAlg} is defined as follows:
\begin{lstlisting}
trait SubAlg[E] extends ExpAlg[E] {
    def sub(e1 : E, e2 : E) : E
}
\end{lstlisting}
Now evaluation on the new language can be realized by code reuse, and without modifying existing code:
\begin{lstlisting}
trait SubEval extends Eval {
    def sub(e1 : Int, e2 : Int) = e1 - e2
}
\end{lstlisting}
On this, to create an expression of \lstinline{SubAlg}, a generic method is defined as follows:
\begin{lstlisting}
def build[E](alg : SubAlg[E]) : E =
    alg.sub(alg.add(alg.lit(2), alg.lit(3)), alg.lit(4))
\end{lstlisting}
Such a method implicitly represents the expression \lstinline{"2 + 3 - 4"}. The code
\begin{lstlisting}
build(new Eval(){})
\end{lstlisting}
results in \lstinline{1}, as the result of evaluation. Now we observe that an ``object'' of expression actually
has the function type \lstinline{SubAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to refactor an \lstinline{ExpAlg} expression to a \lstinline{SubAlg} expression by
replacing all negative numbers with subtraction, a transformation algebra can be implemented as follows:
\begin{lstlisting}
trait Refactor[E] extends ExpAlg[E] {
    def alg : SubAlg[E]
    def lit(i : Int) = if (i > 0) alg.lit(i) else alg.sub(alg.lit(0), alg.lit(-i))
    def add(e1 : E, e2 : E) = alg.add(e1, e2)
}
\end{lstlisting}
With \lstinline{Refactor}, the expression \lstinline{"1 + (-2)"} is refactored to \lstinline{"1 + (0 - 2)"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply refactoring to an \lstinline{ExpAlg} expression before \lstinline{alg} returns the result.

### 2016-10-15 paper/sections/Library/OpenRecursion.tex
More text.

In this section, we will further look into the mechanism behind our modular parsing library.
We will first introduce some concepts independently from the process of parsing, including open recursion and Object Algebras,
then present how they are integrated in our library.
Starting from simple examples, a series of extensions and refactorings will be applied to illustrate
how we achieve modularity in a type-safe way.

@@

Open recursion is known as a useful feature that one method body can ``invoke another method of the same object via a special variable called \lstinline{self} or, in some languages, \lstinline[keywords={}]{this}'' (by Ralf Hinze). The interesting thing is that such a variable \lstinline{self} is late-bound, or open to the recursion, which means it can integrate some additional features defined later, whereas the method that takes \lstinline{self} as a parameter can simply delegate the known cases to it.

For sure we cannot always leave it open; upon using it we have to close the recursion, at which time we need to take a fix-point. Here one might suddenly be reminded of the \lstinline{fix} function in some functional languages like Haskell. Below is the famous Fibonacci example:
\begin{lstlisting}[language=Haskell,keywords={}]
fib :: Int -> Int
fib 0 = 0
fib 1 = 1
fib n = fib (n - 1) + fib (n - 2)
\end{lstlisting}
An alternative approach using open recursion is as follows:
\begin{lstlisting}[language=Haskell,keywords={type}]
type Fix t = t -> t

fib' :: Fix (Int -> Int)
fib' self 0 = 0
fib' self 1 = 1
fib' self n = self (n - 1) + self (n - 2)
\end{lstlisting}
where \lstinline{self} is the explicit self-reference, namely the delegator. Functions like \lstinline{fib'} which takes
an explicit self-reference as its parameter are called \textit{generators} (by Ralf Hinze). The definition of \lstinline{fix} is:
\begin{lstlisting}[language=Haskell,keywords={}]
fix f = let x = f x in x
\end{lstlisting}
If we close the recursion on \lstinline{fib'} only, the evaluation of \lstinline{fix fib'} \lstinline{2} is as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix fib' 2
= let x = fib' x in x 2
= let x = fib' x in fib' x 2
= let x = fib' x in (x 1) + (x 0)
= let x = fib' x in (fib' x 1) + (fib' x 0)
= let x = fib' x in 1 + 0
= 1
\end{lstlisting}
which behaves the same as \lstinline{fib 2}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib'}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.

Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\begin{lstlisting}[language=Haskell,keywords={}]
(*) :: Fix (a -> b) -> Fix (a -> c) -> Fix (a -> (b, c))
f * g = \self x -> (f (fst . self) x, g (snd . self) x)
\end{lstlisting}
Then we define a generator which prints out a number as a string:
\begin{lstlisting}[language=Haskell,keywords={}]
show' :: Fix (Int -> String)
show' self x = show x
\end{lstlisting}
Now \lstinline{fix (fib'}\lstinline{ * show')} \lstinline{2} results in \lstinline{(1, "2")}. We have seen that both \lstinline{fib'} and \lstinline{show'} are integrated in the evaluation. In Scala, however, there is no \lstinline{fix} function, but it can be easily defined since it is based on lazy evaluation. See below:
\begin{lstlisting}
def fix[A](f : (=> A) => A) : A = {
    lazy val a : A = f(a)
    a
}
\end{lstlisting}
Next we define the general combinator \lstinline{merge} for open recursion. It is similar to the \lstinline{(*)} above, but instead of putting two values in a pair by default, \lstinline{merge} takes a parameter called \lstinline{op}, which tells the associative operation of two results. The definition is straightforward:
\begin{lstlisting}
def merge[E, F, G, H[_,_]](op : F => G => H[F, G], x : (=> E) => F, y : (=> E) => G) : (=> E) => H[F, G] = e => op(x(e))(y(e))
\end{lstlisting}

>>>>

In this section, we will further look into the mechanism behind our modular parsing library.
We will first introduce some concepts independently from the process of parsing, including open recursion and Object Algebras,
then present how they are integrated in our library.
Starting from simple examples, a series of extensions and refactoring will be applied to illustrate
how we achieve modularity in a type-safe way.

@@

Open recursion is known as a useful feature that one method body can ``invoke another method of the same object via a special variable called \lstinline{self} or, in some languages, \lstinline[keywords={}]{this}'' (by Ralf Hinze). The interesting thing is that such a variable \lstinline{self} is late-bound, or open to the recursion, which means it can integrate some additional features defined later, while existing methods that take \lstinline{self} as a parameter can simply delegate their known cases to it.

For sure we cannot always leave it open; upon using it we have to close the recursion, at which time we need to take a fix-point. Here one might suddenly be reminded of the \lstinline{fix} function in some functional languages like Haskell. Below is the famous Fibonacci example:
\begin{lstlisting}[language=Haskell,keywords={}]
fib :: Int -> Int
fib 0 = 0
fib 1 = 1
fib n = fib (n - 1) + fib (n - 2)
\end{lstlisting}
An alternative implementation using open recursion is as follows:
\begin{lstlisting}[language=Haskell,keywords={type}]
type Fix t = t -> t

fib' :: Fix (Int -> Int)
fib' self 0 = 0
fib' self 1 = 1
fib' self n = self (n - 1) + self (n - 2)
\end{lstlisting}
where \lstinline{self} is the explicit self-reference, namely the delegator. Functions like \lstinline{fib'} which takes
an explicit self-reference as its parameter are called \textit{generators} (by Ralf Hinze). The definition of \lstinline{fix} is:
\begin{lstlisting}[language=Haskell,keywords={}]
fix f = let x = f x in x
\end{lstlisting}
If we close the recursion on \lstinline{fib'} only, the evaluation of \lstinline{fix fib'} \lstinline{2} is processed as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix fib' 2
= let x = fib' x in x 2
= let x = fib' x in fib' x 2
= let x = fib' x in (x 1) + (x 0)
= let x = fib' x in (fib' x 1) + (fib' x 0)
= let x = fib' x in 1 + 0
= 1
\end{lstlisting}
which behaves the same as \lstinline{fib 2}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib'}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.

Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\begin{lstlisting}[language=Haskell,keywords={}]
(*) :: Fix (a -> b) -> Fix (a -> c) -> Fix (a -> (b, c))
f * g = \self x -> (f (fst . self) x, g (snd . self) x)
\end{lstlisting}
Then we define a generator which prints out a number as a string:
\begin{lstlisting}[language=Haskell,keywords={}]
show' :: Fix (Int -> String)
show' self x = show x
\end{lstlisting}
Now \lstinline{fix (fib'}\lstinline{ * show')} \lstinline{2} results in \lstinline{(1, "2")}. We have seen that both \lstinline{fib'} and \lstinline{show'} are integrated in the evaluation. In Scala, however, there is no library \lstinline{fix} function, but it can be easily defined since it is based on lazy evaluation. See below:
\begin{lstlisting}
def fix[A](f : (=> A) => A) : A = {
    lazy val a : A = f(a)
    a
}
\end{lstlisting}
Next we define the general combinator \lstinline{merge} for open recursion. It is similar to the \lstinline{(*)} above, but instead of putting two values in a pair by default, \lstinline{merge} takes a parameter called \lstinline{op}, which tells the associative operation of two results. The definition is straightforward:
\begin{lstlisting}
def merge[E, F, G, H[_,_]](op : F => G => H[F, G], x : (=> E) => F, y : (=> E) => G) : (=> E) => H[F, G] = e => op(x(e))(y(e))
\end{lstlisting}

### 2016-10-15 paper/sections/Library/ParsingLibrary.tex
More text.

>>>>

In the previous sections we have talked about open recursion and Object Algebras. These concepts will later be composed together for our
initial goal, which is to build extensible parsers, but surely it cannot live without a parsing library. But we want to argue here that besides
open recursion, which constructs the basis of extensibility in parsing, the other aspects are totally orthogonal, and hence can have their alternatives.
For example, Object Algebras are helpful to build extensible data structures, whereas an alternative with the same function can still be used. And a better
parsing library would introduce a lot of fancy features, nevertheless, it is independent from our key idea.

In Scala we select Packrat Parsing for our prototype implementation. It is practical not only because of the performance, and its parser combinators, but also due to its support for direct left-recursion. From theoretical point of view, the algorithm behind Packrat parsers supports both direct and indirect left-recursion. The current version of the library is still buggy in some cases with indirect left-recursion, but we believe that they will fix it in the future, and with direct left-recursion it is already practical to use.

Moreover, the Scala parsing library provides us with a number of useful parser combinators, with which the syntax for parsing looks concise and clear. Besides \lstinline{|} for alternation and \lstinline{~} for sequential composition, we also have \lstinline{rep} and \lstinline{repsep} for repetitions, \lstinline{chainl1} for left-associative grammar rules, and so on. Specifically, a new combinator \lstinline{|||} is available, also for alternative but with longest match composition. Imagine that we are composing a list of parsers using alternative \lstinline{|}, in an order. A parser in the front might be successful in parsing a substring of the input, but then it unexpectedly stops the parsing and discards the rest, while another parser might be able to parse the whole input. In that case, using \lstinline{|||} we no longer need to worry about the order of composition, which encourages its extensibility. A concrete example will be illustrated for this later.

Generally, a Packrat parser has the type \lstinline{PackratParser[E]} for some \lstinline{E}. Whereas our modular parser has a different type, which implies that we will not apply \lstinline{|||} for composition directly, but use \lstinline{|||} to implement a new combinator for convenience. It will also be introduced afterwards.
